WSClean: an implementation of a fast, generic wide-field imager

### 2014-02-25 articles/2014/Article-WSClean/wsclean.tex
Small changes

\subsection{Discretization of $w$} \label{sec:gridding-w}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/aliasing-example-annotated}
\caption{Aliasing artefacts caused by insufficient $w$-layers in a simulated field. WSClean was set to use 12 $w$-layers. The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers. Sources are 1~Jy (red circles), ghost sources are approximately 0.2~Jy. Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce more ghost sources. }
\label{fig:aliasing-example}
\end{center}
\end{figure}
While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of $w$-layers that need to be processed. For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian. This results in the constraint
\begin{equation} \label{eq:minimum-w-distance}
\left|\left(w_A - w_B\right) 2\pi (\sqrt{1-l^2-m^2}-1)\right| \ll 1.
\end{equation}
This suggest that a uniform discretization in $w$ is optimal (TODO this is in contrast to what Cornwell 2008 says -- Cornwell mailed me he made a mistake... Do we need to explicitly point that out here :-/ ?). From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
\begin{equation} \label{eq:nwlayers-bound}
 N_\textrm{wlayers} \gg 2\pi \left(w_{\max} - w_{\min}\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
\end{equation}
Actual values for the right-hand side can differ a lot. The value of $w_{\max} - w_{\min}$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image. For the MWA, a typical value of $w_{\max} - w_{\min}$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$. For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$. This implies that tens of $w$-layers are required at zenith and hundreds at lower elevations. The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in the next sections.

@@

\subsection{Beam correction}
\begin{figure*}
\begin{center}
\includegraphics[width=6cm]{img/pulsar-stokesi}\hspace{1cm}\includegraphics[width=6cm]{img/pulsar-stokesv}
\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with WSClean. The pulsar (centre of image) displays 17\% circular polarization.}
\label{fig:pulsar-stokes-iv}
\end{center}
\end{figure*}
Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such. However, this is incorrect because i) $X$ is normally defined as being parallel to the horizon; and ii) ($2\times$) Stokes I$\neq pp + qq$.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. Each polarization requires a separate run of the $w$-stacking algorithm. Once these six images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. The method is demonstrated in Fig.~\ref{fig:pulsar-stokes-iv}, which displays a detection of pulsar J0437-4715 in Stokes V. The image was made with an initial pipeline for the radio-sky monitor project that uses the MWA to search for transients, including exoplanets (TODO cite). The pipeline includes WSClean to make the Stokes-parameter images.

Our image-based beam correction method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam. To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm excessively unless an a-projection kernel is used. The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.

@@

Table~\ref{tbl:accuracy-measurements} lists the measured RMS in the residual image and the standard errors of the source brightnesses as detected by Aegean. In all tests, WSClean performs slightly more accurate. At zenith, CASA produces a residual RMS twice as high as WSClean. This is caused by the fact that we keep the number of $w$-projection planes in $w$-projection equal to the number of $w$-layers in $w$-stacking, resulting in only 12 $w$-planes at zenith. This evidently has a stronger effect on the $w$-projection algorithm. However, the computational performance of the $w$-projection algorithm is hardly affected by the number of $w$-projection planes, and in practical situations one would always use more $w$-projection planes. When 128 $w$-projection planes are used in CASA, its residual RMS is equal to WSClean with 12 $w$-layers, but the flux density measurements are less accurate. We do not know why this parameter needs to be higher in CASA to reach the same RMS (TODO: do I want to find out this is because of invalid sqrt(w) gridding?).  Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlayers}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree. Also unexpected is that the source flux density becomes worse by increasing the number of planes. For WSClean both values stay approximately the same when the number of $w$-layers is increased.

>>>>

\subsection{Discretization of $w$} \label{sec:gridding-w}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/aliasing-example-annotated}
\caption{Aliasing artefacts caused by insufficient $w$-layers in a simulated field. WSClean was set to use 12 $w$-layers. The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers. Sources are 1~Jy (red circles), ghost sources are approximately 0.2~Jy. Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce more ghost sources. }
\label{fig:aliasing-example}
\end{center}
\end{figure}
While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of $w$-layers that need to be processed. For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian. This results in the constraint
\begin{equation} \label{eq:minimum-w-distance}
\left|\left(w_A - w_B\right) 2\pi (\sqrt{1-l^2-m^2}-1)\right| \ll 1.
\end{equation}
This suggest that a uniform discretization in $w$ is optimal. This is in contrast to \citet{wprojection-cornwell} where $\sqrt{w}$ tabulation is suggested. From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
\begin{equation} \label{eq:nwlayers-bound}
 N_\textrm{wlayers} \gg 2\pi \left(w_{\max} - w_{\min}\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
\end{equation}
Actual values for the right-hand side can differ a lot. The value of $w_{\max} - w_{\min}$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image. For the MWA, a typical value of $w_{\max} - w_{\min}$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$. For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$. This implies that tens of $w$-layers are required at zenith and hundreds at lower elevations. The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in the next sections.

@@

\subsection{Beam correction}
\begin{figure*}
\begin{center}
\includegraphics[width=6cm]{img/pulsar-stokesi}\hspace{1cm}\includegraphics[width=6cm]{img/pulsar-stokesv}
\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with WSClean. The pulsar (centre of image) displays 17\% circular polarization.}
\label{fig:pulsar-stokes-iv}
\end{center}
\end{figure*}
Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such. However, this is incorrect because they are not necessarily orthogonal, i.e., ($2\times$) Stokes I$\neq pp + qq$.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. This requires four runs of the algorithm. Once these images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. The method is demonstrated in Fig.~\ref{fig:pulsar-stokes-iv}, which displays a detection of pulsar J0437-4715 in Stokes V. The image was made with an initial pipeline for the radio-sky monitor project that uses the MWA to search for transients and exoplanets. The pipeline includes WSClean to make the Stokes-parameter images.

Our image-based beam-correction method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam. To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm excessively unless an a-projection kernel is used. The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.

@@

Table~\ref{tbl:accuracy-measurements} lists the measured RMS in the residual image and the standard errors of the source brightnesses as detected by Aegean. In all tests, WSClean performs slightly more accurate. At zenith, CASA produces a residual RMS twice as high as WSClean. This is caused by the fact that we keep the number of $w$-projection planes in $w$-projection equal to the number of $w$-layers in $w$-stacking, resulting in only 12 $w$-planes at zenith. This evidently has a stronger effect on the $w$-projection algorithm. However, the computational performance of the $w$-projection algorithm is hardly affected by the number of $w$-projection planes, and in practical situations one would always use more $w$-projection planes. When 128 $w$-projection planes are used in CASA, its residual RMS is equal to WSClean with 12 $w$-layers, but the flux density measurements are less accurate. We do not know why this parameter needs to be higher in CASA to reach the same RMS.  Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlayers}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree. Also unexpected is that the source flux density becomes worse by increasing the number of planes. For WSClean both values stay approximately the same when the number of $w$-layers is increased.
